---
title: "WHO Moratlity Data"
author: "S Bashir"
date: "9 October 2016"
output: html_document
---

In this document, the steps to *tidy* the [World Health Organisation's (WHO)](www.who.int) mortality data for use in R are documented. The following steps are presented.

* Reading in the WHO mortality data
* Keeping the **neoplasm**, **melanoma**, **lymphoma** and **leukaemia** data for ICD10 part 1
* Reading in the population data
* Merging the population data to the neoplasm data
  + Care has to be taken with respect to the format of the age groups. 
  + Combine the youngest age groups into a new 0-4 years. 
  + Transform the neoplasm and population datasets from wide to long before merging.
  
* Calculating the mortality rate per 100,000 population in a new variable
  + rate = (100000*deaths)/pop
* Writing a function that will extract data based on country, cause of death, year and sex
  + If any of these are left blank then all the data is extracted
  + A data frame containing the requested data should be returned.
  
* Next steps will be defined later
  * Reading the International Classificaion of Diseases data
  * Sumamry graphs
  * Summary tables
  
**Note** There is no not need to use *R Shiny* for this exercise. Only program in this R markdown document.

## Data source

The data was downloaded from <http://www.who.int/healthinfo/mortality_data/en/>. It is assumed that data has been downloaded and resides in a directory "OrigData". The required files are:

* MortIcd7
* MortIcd8
* MortIcd9
* MortIcd10_part1
* MontIcd10_part2
* pop


## Downloading libraries

```{r DownloadLibraries, echo=FALSE}
library(stringr)
library(readr)
library(tibble)
library(plyr)
library(xml2)
```

## Reading in the WHO mortality data

The following code will automatically read in the WHO mortality data. For now, we will read only the Morticd10_part1.

```{r ReadWHOdata, echo=FALSE}

# Set the directory where you have saved the data. 
# This is not a great solution but it works and is easier for the moment, Later it should change.
setwd("OrigData/")

# Select the data file
csvWHO <- "Morticd10_part1"

# If we wanted to read all data files( excluding .zip, .doc, .xls)
# csvWHO <- list.files(path=".", full.names = FALSE) %>% str_subset("^((?!zip|doc|xls).)*$")

# Use the lapply to read all the data into one list
dataList <- lapply(csvWHO, read_csv, guess_max = min(300000, Inf))

# Extract the data into data.frames (tbl) but start by giving them the dataest you want.
# I convert them to lower case as they are not consistent in the original.
names(dataList) <- str_to_lower(csvWHO) 
list2env(dataList, envir=.GlobalEnv)

# dataList is not longer needed so we will remove it.
rm(dataList)

# Removing the ICD 7, 8, 9 and 10 part 2 data as we will not use them for the moment.
# Also we do not want to have problems with the memory 
# rm(morticd7, morticd8, morticd9, morticd10_part2) # Only necessary when we read all data files.

# Some garbage collection to free up the memory (although this is something that is
# automatically by kernel). In future we may not need this step.
gc()
```

## Renaming columns

Columns Deaths1,...,Deaths26 are given more more descriptive names. These column names will also be used in the population dataset to allow merging the two.

```{r renameAgeCols_in_Deaths}
renamed_morticd10_part1 <- morticd10_part1 %>%
  rename(c("Deaths1"="AllAges")) %>%
  rename(c("Deaths2"="Age0")) %>%
  rename(c("Deaths3"="Age1")) %>%
  rename(c("Deaths4"="Age2")) %>%
  rename(c("Deaths5"="Age3")) %>%
  rename(c("Deaths6"="Age4")) %>%
  rename(c("Deaths7"="Age5to9")) %>%
  rename(c("Deaths8"="Age10to14")) %>%
  rename(c("Deaths9"="Age15to19")) %>%
  rename(c("Deaths10"="Age20to24")) %>%
  rename(c("Deaths11"="Age25to29")) %>%
  rename(c("Deaths12"="Age30to34")) %>%
  rename(c("Deaths13"="Age35to39")) %>%
  rename(c("Deaths14"="Age40to44")) %>%
  rename(c("Deaths15"="Age45to49")) %>%
  rename(c("Deaths16"="Age50to54")) %>%
  rename(c("Deaths17"="Age55to59")) %>%
  rename(c("Deaths18"="Age60to64")) %>%
  rename(c("Deaths19"="Age65to69")) %>%
  rename(c("Deaths20"="Age70to74")) %>%
  rename(c("Deaths21"="Age75to79")) %>%
  rename(c("Deaths22"="Age80to84")) %>%
  rename(c("Deaths23"="Age85to89")) %>%
  rename(c("Deaths24"="Age90to94")) %>%
  rename(c("Deaths25"="Age95More")) %>%
  rename(c("Deaths26"="AgeUnspecified"))

head(renamed_morticd10_part1)
```

## 3)	Drop columns IM_Frmat, IM_Deaths1…, IM_Deaths4

For now, we won't go to the detail given by these age groups. 

```{r DropAgeColumns_in_Deaths}
reduced_morticd10 <- select(renamed_morticd10_part1, -IM_Frmat, -IM_Deaths1, -IM_Deaths2, -IM_Deaths3, -IM_Deaths4)
head(reduced_morticd10)
```

## Create new column Age0to4 in the deaths dataset

In order to create a new column with the deaths for ages 0 to 4, it is useful to transform the NA's in 0 and then simply to create a new column equal to the sum of columns Age0, Age1, Age2, Age3 and Age4.

```{r NewColumnAge0to4_in_Deaths}
# Replace NA's in these columns by 0's
reduced_morticd10[c("Age0","Age1","Age2","Age3","Age4")][is.na(reduced_morticd10[c("Age0","Age1","Age2","Age3","Age4")])] <- 0

# Create new column Age0to4
reduced_morticd10$Age0to4 <- reduced_morticd10$Age0 + reduced_morticd10$Age1 + reduced_morticd10$Age2 + reduced_morticd10$Age3 + reduced_morticd10$Age4

reduced_morticd10
```

## 6)	Keep neoplasms, leukaemias, melanomas and lymphoma. Drop columns Cause.

Once we create new datasets only for a specific disease, we can drop the column Cause.

```{r keepNeoplasm}
# First, keep only Neoplasm
neoplasm_deaths <- reduced_morticd10 %>% filter(Cause == 1026) %>% select(-Cause)
filter(neoplasm_deaths,List!=101) # Note that all records have List == 101
```

```{r keepMelanoma}
# First, keep only Neoplasm
melanoma_deaths <- reduced_morticd10 %>% filter(Cause == 1035) %>% select(-Cause)
filter(melanoma_deaths,List!=101) # All records have List == 101
```

```{r keepLymphoma}
# First, keep only Neoplasm
lymphoma_deaths <- reduced_morticd10 %>% filter(Cause == 1043) %>% select(-Cause)
filter(lymphoma_deaths,List!=101) # All records, List == 101
```

```{r keepLeukaemia}
# First, keep only Neoplasm
leukaemia_deaths <- reduced_morticd10 %>% filter(Cause == 1045) %>% select(-Cause)
filter(leukaemia_deaths,List!=101)
```

## Drop column List in deaths dataset

By observing leukaemia_deaths, melanoma_deaths, neoplasm_deaths and lymphoma_deaths, we can see that all their records have List == 101. Therefore, we can simply drop this column.

```{r DropListColumn}
neoplasm_deaths <- select(neoplasm_deaths,-List)
lymphoma_deaths <- select(lymphoma_deaths,-List)
melanoma_deaths <- select(melanoma_deaths,-List)
leukaemia_deaths<- select(leukaemia_deaths,-List)
```

## Transform Deaths from wide to long (neoplasm dataset)

This is done by grouping columns Age0, ..., Age26 into 2 new columns AgeGroup and NDeaths (number of deaths).

```{r WideToLongNeoplasm}
# Gather columns Deaths1, ..., IM_Deaths4 into 2 new columns AgeGroup and NDeaths
long_neoplasm_deaths <- neoplasm_deaths %>%
  gather(AgeGroup, NDeaths, AllAges:Age0to4)

head(long_neoplasm_deaths)
```


## Read the population dataset

```{r ReadPop}
# Reading the csv file
csvPOP <- "pop"
dataList <- lapply(csvPOP, read_csv, guess_max = min(300000, Inf))
names(dataList) <- str_to_lower(csvPOP) 
list2env(dataList, envir=.GlobalEnv) 
```

## Change the age column names in population dataset

In order to match with the column names in the deaths datasets and allow merging, we need to rename the columns in the population dataset too.

```{r RenameColsPop}
renamed_pop <- pop %>%
  rename(c("Pop1"="AllAges")) %>%
  rename(c("Pop2"="Age0")) %>%
  rename(c("Pop3"="Age1")) %>%
  rename(c("Pop4"="Age2")) %>%
  rename(c("Pop5"="Age3")) %>%
  rename(c("Pop6"="Age4")) %>%
  rename(c("Pop7"="Age5to9")) %>%
  rename(c("Pop8"="Age10to14")) %>%
  rename(c("Pop9"="Age15to19")) %>%
  rename(c("Pop10"="Age20to24")) %>%
  rename(c("Pop11"="Age25to29")) %>%
  rename(c("Pop12"="Age30to34")) %>%
  rename(c("Pop13"="Age35to39")) %>%
  rename(c("Pop14"="Age40to44")) %>%
  rename(c("Pop15"="Age45to49")) %>%
  rename(c("Pop16"="Age50to54")) %>%
  rename(c("Pop17"="Age55to59")) %>%
  rename(c("Pop18"="Age60to64")) %>%
  rename(c("Pop19"="Age65to69")) %>%
  rename(c("Pop20"="Age70to74")) %>%
  rename(c("Pop21"="Age75to79")) %>%
  rename(c("Pop22"="Age80to84")) %>%
  rename(c("Pop23"="Age85to89")) %>%
  rename(c("Pop24"="Age90to94")) %>%
  rename(c("Pop25"="Age95More")) %>%
  rename(c("Pop26"="AgeUnspecified"))


head(renamed_pop)
```

## Create new column Age0to4 in population dataset

In order to create a new column with the deaths for ages 0 to 4, it is useful to transform the NA's in 0 and then simply to create a new column equal to the sum of columns Age0, Age1, Age2, Age3 and Age4.

```{r NewAge0to4_in_Pop}
# Replace NA's in these columns by 0's
renamed_pop[c("Age0","Age1","Age2","Age3","Age4")][is.na(renamed_pop[c("Age0","Age1","Age2","Age3","Age4")])] <- 0

# Create new column Age0to4
renamed_pop$Age0to4 <- renamed_pop$Age0 + renamed_pop$Age1 + renamed_pop$Age2 + renamed_pop$Age3 + renamed_pop$Age4
head(renamed_pop)
```

## Drop Lb (Live birth) column in population dataset

We will disregards this information for now.

```{r DropLbPop}
renamed_pop <- select(renamed_pop,-Lb)
head(renamed_pop)
```

## Transform population dataset from wide to long

```{r PopWideToLong}
long_population <- renamed_pop %>%
  gather(AgeGroup, NPop, AllAges:Age0to4) # Gather Pop1, ..., Pop26 and Lb into 2 new columns

head(long_population)
```

### Merging the population data to the neoplasm mortality data

* Care has to be taken with respect to the format of the age groups. 
  + The naming for age groups has been made uniform for the 2 datasets.
* Combine the youngest age groups into a new 0-4 years. 
  + A new age group for 0-4 years has been created in the 2 datasets.
* Consider adding an **all** age group category (**not urgent**)
  + This age group already existed in the original dataset.
* Calculate the mortality rate per 100,000 population in a new variable
  + rate = (100000*deaths)/pop

```{r MergeDeathsAndPop}
# It is to be noted that long_population has records for the years 1950-2015 while long_neoplasm_deaths has only the years 1996-2004.
neoplasm_deathsAndPop <- merge(long_neoplasm_deaths,long_population)
neoplasm_deathsAndPop
```

Before calculating the mortality rate, we need to eliminate records with NPop equal to 0 or NA. Also we need to transform all NDeaths=NA to NDeaths=0.

```{r Clear_DeathsAndPop}
# Filter zero or NA populations
neoplasm_deathsAndPop <- neoplasm_deathsAndPop %>%
  filter(NPop != 0) %>% 
  filter(!is.na(NPop))

# Transform NA deaths to zero deaths
neoplasm_deathsAndPop$NDeaths[is.na(neoplasm_deathsAndPop$NDeaths)] <- 0

neoplasm_deathsAndPop
```

Create new column with the mortality rate per 100000:
```{r CalcMortalityRate}
neoplasm_deathsAndPop$MortRate <- 100000 * neoplasm_deathsAndPop$NDeaths/neoplasm_deathsAndPop$NPop
head(neoplasm_deathsAndPop)
```




## Write a function that will extract data based on country, cause of death, year and sex
  * If any of these are left blank then all the data is extracted
  * A data frame containing the requested data should be returned.

```{r}
getWHOdata <- function(country, causes, year, sex){
  # country - Vector of countries to be filtered
  #   cause - Vector of causes of death to be filtered
  #    year - Vector of years to be filtered
  #     sex - Vector of gender to be filtered

}


```


### Next steps

Once the above steps have been completed, the next steps will be defined. They will include:
  * Reading the International Classificaion of Diseases data
  * Sumamry graphs
  * Summary tables
  

## Reading the International Classificaion of Diseases data

**Joana:** Please read in the data form the word document here.

```{r readICDdata, echo=TRUE}
library(xml2)

```
